---
title: "Exploration"
author: "Student"
date: today
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(Lahman)
library(crosstalk)
library(plotly)
library(DT)
```

The Exploration tab allows you to investigate the physical profiles of MLB players across defensive positions during the 2023 season.

### Position Averages

The table below summarizes the average height, average weight, and active player count by position.

```{r}
#| label: position-averages
#| echo: false
#| message: false
#| warning: false

# Primary position for each player
player_pos <- Fielding |>
  filter(yearID == 2023) |>
  group_by(playerID, POS) |>
  summarize(games = sum(G), .groups = "drop") |>
  group_by(playerID) |>
  slice_max(games, n = 1, with_ties = FALSE) |>
  ungroup()

mlb_2023 <- player_pos |>
  inner_join(People, by = "playerID") |>
  filter(!is.na(height), !is.na(weight)) |>
  mutate(
    Position = case_when(
      POS == "P"  ~ "Pitcher",
      POS == "C"  ~ "Catcher",
      POS == "1B" ~ "First Base",
      POS == "2B" ~ "Second Base",
      POS == "3B" ~ "Third Base",
      POS == "SS" ~ "Shortstop",
      POS == "OF" ~ "Outfield",
      TRUE ~ POS
    ),
    Player = paste(nameFirst, nameLast)
  )

# Compute averages by position
pos_summary <- mlb_2023 |>
  group_by(Position) |>
  summarize(
    Players = n(),
    `Avg Height (in)` = round(mean(height), 1),
    `Avg Weight (lbs)` = round(mean(weight), 1),
    .groups = "drop"
  ) |>
  arrange(desc(`Avg Weight (lbs)`))

datatable(
  pos_summary,
  rownames = FALSE,
  options = list(pageLength = 10, dom = "t")
)
```

### Interactive Height vs. Weight Scatter

Use the dropdown filter below to pick a specific position or view all players simultaneously. You can also hover over individual points to inspect each athlete's details, or double-click items in the legend.

```{r}
#| label: interactive-scatter
#| echo: false
#| warning: false
#| message: false

sd <- SharedData$new(mlb_2023)

filter_select(
  id = "pos_filter",
  label = "Select Position to Filter:",
  sharedData = sd,
  group = ~Position,
  multiple = TRUE
)

plot_ly(
  sd,
  x = ~height,
  y = ~weight,
  color = ~Position,
  text = ~paste0("Player: ", Player, "<br>Position: ", Position, "<br>Height: ", height, " in<br>Weight: ", weight, " lbs"),
  hoverinfo = "text",
  type = "scatter",
  mode = "markers",
  marker = list(size = 7, opacity = 0.7)
) |>
  layout(
    title = "Interactive MLB Player Height & Weight (2023)",
    xaxis = list(title = "Height (inches)"),
    yaxis = list(title = "Weight (pounds)"),
    legend = list(title = list(text = "<b>Position</b>"))
  )
```
