---
title: "Height and Weight in Major League Baseball"
author: "Student"
date: today
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(Lahman)
```

In Major League Baseball, physical stature varies considerably across different roles on the diamond. A first baseman or pitcher often brings height and power, while middle infielders lean toward agility. 

The scatterplot below visualizes player height and weight for the 2023 MLB season (~1,500 active players), colored by defensive position.

```{r}
#| label: height-weight-scatter
#| echo: false
#| warning: false
#| message: false
#| fig-width: 9
#| fig-height: 6

# Determine primary position for each player in 2023
player_pos <- Fielding |>
  filter(yearID == 2023) |>
  group_by(playerID, POS) |>
  summarize(games = sum(G), .groups = "drop") |>
  group_by(playerID) |>
  slice_max(games, n = 1, with_ties = FALSE) |>
  ungroup()

# Join with People table for height and weight
mlb_2023 <- player_pos |>
  inner_join(People, by = "playerID") |>
  filter(!is.na(height), !is.na(weight)) |>
  mutate(
    Position = case_when(
      POS == "P"  ~ "Pitcher",
      POS == "C"  ~ "Catcher",
      POS == "1B" ~ "First Base",
      POS == "2B" ~ "Second Base",
      POS == "3B" ~ "Third Base",
      POS == "SS" ~ "Shortstop",
      POS == "OF" ~ "Outfield",
      TRUE ~ POS
    )
  )

# Scatterplot of height vs weight
ggplot(mlb_2023, aes(x = height, y = weight, color = Position)) +
  geom_jitter(alpha = 0.65, size = 2, width = 0.25, height = 0.5) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "MLB Player Dimensions (2023 Season)",
    subtitle = "Height and weight across 1,400+ players, colored by primary defensive position",
    x = "Height (inches)",
    y = "Weight (pounds)",
    color = "Position",
    caption = "Source: Lahman Baseball Database"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
```

A slight positive relationship exists between height and weight overall. Pitchers and first basemen cluster toward the upper-right corner (taller and heavier), whereas second basemen and shortstops concentrate toward the lower-left.
