---
title: "Historical Growth Simulator"
author: "Student"
date: today
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(Lahman)
library(plotly)
library(DT)
```

This simulator animates the physical evolution of Major League Baseball players across a century of baseball history (1920–2020). By keeping the axes constant, you can press **Play** (or drag the slider) to watch thousands of players steadily expand in both height and weight over time.

### Animated Decade-by-Decade Scatter

Click **Play** below or drag the year slider to simulate the passage of time across 11 distinct decades. Notice how the central mass of players drifts toward the upper-right quadrant.

```{r}
#| label: animated-simulator
#| echo: false
#| warning: false
#| message: false

# Sample decade snapshots from 1920 to 2020
decades <- seq(1920, 2020, by = 10)

sim_data <- Appearances |>
  filter(yearID %in% decades) |>
  distinct(playerID, yearID) |>
  inner_join(People, by = "playerID") |>
  filter(!is.na(height), !is.na(weight)) |>
  mutate(
    Year = factor(yearID),
    Player = paste(nameFirst, nameLast)
  )

plot_ly(
  sim_data,
  x = ~height,
  y = ~weight,
  frame = ~Year,
  text = ~paste0("<b>", Player, "</b><br>Height: ", height, " in<br>Weight: ", weight, " lbs"),
  hoverinfo = "text",
  type = "scatter",
  mode = "markers",
  marker = list(size = 6, color = "#2c3e50", opacity = 0.55)
) |>
  layout(
    title = "MLB Player Dimensions (Decade Snapshots: 1920–2020)",
    xaxis = list(title = "Height (inches)", range = c(64, 84)),
    yaxis = list(title = "Weight (pounds)", range = c(130, 290)),
    margin = list(t = 50, b = 50)
  ) |>
  animation_opts(
    frame = 900,
    transition = 450,
    easing = "cubic-in-out"
  ) |>
  animation_slider(
    currentvalue = list(prefix = "Active Season: ", font = list(color = "#2c3e50", size = 14))
  )
```

### Decade Summary Table

The table below calculates the exact average height, average weight, and Body Mass Index (BMI) for each simulated decade.

```{r}
#| label: decade-summary
#| echo: false
#| warning: false
#| message: false

decade_table <- sim_data |>
  group_by(Decade = Year) |>
  summarize(
    `Active Players` = n(),
    `Avg Height (in)` = round(mean(height), 1),
    `Avg Weight (lbs)` = round(mean(weight), 1),
    `Avg BMI` = round(mean((weight / (height^2)) * 703), 1),
    .groups = "drop"
  )

datatable(
  decade_table,
  rownames = FALSE,
  options = list(pageLength = 11, dom = "t")
)
```

### Average Trajectory Over Time

```{r}
#| label: trajectory-plot
#| echo: false
#| warning: false
#| message: false
#| fig-width: 9
#| fig-height: 4.5

p_trend <- decade_table |>
  mutate(Year = as.numeric(as.character(Decade)))

plot_ly(p_trend, x = ~Year) |>
  add_lines(y = ~`Avg Weight (lbs)`, name = "Avg Weight (lbs)", line = list(color = "#e74c3c", width = 3)) |>
  add_lines(y = ~`Avg Height (in)` * 2.5, name = "Avg Height (scaled x2.5)", line = list(color = "#3498db", width = 3, dash = "dash")) |>
  layout(
    title = "Historical Trajectory of Player Averages (1920–2020)",
    xaxis = list(title = "Season"),
    yaxis = list(title = "Value"),
    legend = list(orientation = "h", y = -0.2)
  )
```
